Maximum Entropy Spectral Analysis: a case study

### 2021-06-09 paper/mesa.tex
Response

The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\cite{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

@@

Most importantly, the variational problem admits a closed-form analytical expression for $S(f)$.
The expression was first found by ~\citet{burg1975maximum}:
\begin{equation}\label{eq:MESApsd}
    S(f) = \frac{P_N \Delta t}{\left(\sum_{s=1}^N a_s z^s\right)\left(\sum_{s = 1}^N a^*_s z^{-s}\right)}, 
\end{equation}
where $\Delta t$ is the sampling interval of the time series, $z=\exp{(2\pi i f\Delta t)}$, $a_1$ = 1.
The vector obtained as $(1, a_2, \dots, a_N)$ is also known as the \textit{prediction error filter}.
The coefficients $a_s (s > 0)$, together with an overall multiplicative scale factor $P_N$, are to be determined by an iterative process (called Burg's algorithm). The number $N$ of such coefficients is a choice that shall be made by the user and indeed it is the only hyperparameter that needs to be tuned. The details of the derivation and the actual form for the coefficients $a_s$ can be found in appendix \ref{sec:appendix}.

@@

As noted above, for pushing further the prediction performance from our simple linear predictor, more sophisticated forecasting methods are available in the Machine Learning literature (see e.g. \cite{LSTM_article, oord2016wavenet}) and they can be trained for precision forecasting. However, a Maximum-Entropy trained AR process can be a simple baseline for comparing such more advanced methods and might suffice for many purposes.
This opens a promising path in GW data analysis and other fields of physics might also take advantage from this.

@@

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD. Hence MESA could be employed for {\it online} PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \citep{lalinference}: this scheme assumes {\it stationary data} on a long timescale (longer than the data under study) and it might not reflect the structure of the noise in the analyzed time slice.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \citep{R_ver_2010, R_ver_2011, Edwards_2020, Edwards_2015}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \citep{Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making minimal assumption about the nature of the data under study.
}, making a lighter assumption on the nature of the data. Furthermore, this method would get, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\citep{martini_thesis}.

>>>>

The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\citep{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

@@

Most importantly, the variational problem admits a closed-form analytical expression for $S(f)$.
The expression was first found by ~\citet{burg1975maximum}:
\begin{equation}\label{eq:MESApsd}
    S(f) = \frac{P_N \Delta t}{\left(\sum_{s=0}^{N} a_s z^s\right)\left(\sum_{s = 0}^N a^*_s z^{-s}\right)}, 
\end{equation}
where $\Delta t$ is the sampling interval of the time series, $z=\exp{(2\pi i f\Delta t)}$, $a_0$ = 1.
The vector obtained as $(1, a_1, \dots, a_N)$ is also known as the \textit{prediction error filter}.
The coefficients $a_s (s > 0)$, together with an overall multiplicative scale factor $P_N$, are to be determined by an iterative process (called Burg's algorithm). The number $N$ of such coefficients is a choice that shall be made by the user and indeed it is the only hyperparameter that needs to be tuned. The details of the derivation and the actual form for the coefficients $a_s$ can be found in appendix \ref{sec:appendix}.

@@

As noted above, for pushing further the prediction performance from our simple linear predictor, more sophisticated forecasting methods are available in the Machine Learning literature (see e.g. \citep{LSTM_article, oord2016wavenet, lea2016tcn}) and they can be trained for precision forecasting. However, a Maximum-Entropy trained AR process can be a simple baseline for comparing such more advanced methods and might suffice for many purposes.
This opens a promising path in GW data analysis and other fields of physics might also take advantage from this.

@@

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD. Hence MESA could be employed for {\it online} PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{|\tilde{d}(f)- \tilde{x}(f;\theta)|^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source: this scheme assumes {\it stationary data} on a long timescale (longer than the data under study) and it might not reflect the structure of the noise in the analyzed time slice.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \citep{R_ver_2010, R_ver_2011, Edwards_2020, Chatziioannou_2021}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \citep{Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making minimal assumption about the nature of the data under study.
}, making a lighter assumption on the nature of the data. Furthermore, this method would get, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\citep{martini_thesis}.
Other studies have been done in this direction, mostly using a parametric model for the PSD \citep{Littenberg_2013, Edwards_2015, lalinference}.
